Briefing paper on Ethnicity and COVID-19 status in Scotland

### 2020-07-03 Covid_ethnicity_Scotland_report.Rmd

# Background
There have been reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.   For example a preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95% CI 2.92-5.12); RR 2.11 (95% CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively).

@@

Any additional persons with a registered death with a COVID-19 code ( suspected or probable ) were also identified.

@@

Data were extracted for  this analysis on 15th May 2020. They do not include the Lighthouse data as we do not have access to this through ECOSS as yet.

@@

Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of  recording of self assigned ethnic group in recent years.   Census categories are used. However there remains substantial missingness.   For all cases and controls the linkage team extracted i) the most recent ethnic group assignation from these two datasets  with a 10 year lookback period. Unknown or missing codes were excluded.  Concordance between the most recent SMR ethnic group   and the most frequent  SMR ethnic group  recorded was 98%.

@@

As other sociodemographic factors may vary by ethnicity we further adjusted analyses for residential care home status and Scottish Index of Multiple Deprivation (SIMD).

# Results
SMR based ethnicity was unavailable for approximately 25% of cases and 50% of controls. The ONOMAP approach  assigned ethnicity  in >99% of persons.  Given the substantial and differential missingness of SMR based ethnicity the principle value of showing the  SMR data here is to  inform on the likely absolute  numbers of cases   of non-White ethnicity rather than for the purposes of estimating rate ratios.

@@

Table \ref{tab:crosstab} shows the agreement between the ONOMAP  (name based) assigned ethnicity and the self assigned SMR ethnicity.  The name classification assigns to the same as  SMR category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[5, 5]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.  The proportion with  non-White ethnicity  is low for all groups. Thus  although the sensitivity of ONOMAP for assigning SMR South Asian ethnicity is >99% the positive predictive value of ONOMAP assigned South Asian ethnicity is just 60%.

Clearly analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records.  *Please note that we do not consider the rate ratios here to be reliable because of the differential missingness.* Therefore in all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.

@@

i)all test positive cases ii) all hospitalized or severe  test positive cases, iii) severe  test positive cases iv) severe test positive cases OR a COVID-19 death registration,    Severe test positive cases are those who  entered a critical care unit  or died within 28 days of a positive test.  Note that full follow up to 28 days has not yet completed for  the recent cases. The first column of rate ratios are univariate. Those in the second column are further adjusted for SIMD and care home status.

Subsequent tables show the same comparisons with ethnicity based on ONOMAP name classification. Tables \ref{tab:severity.smr} and \ref{tab:severity} show the distribution over severity categories by ethnicity. Tables \ref{tab:carehome} and \ref{tab:board} show the distribution of care home status and health board.  Note that  p -values  that are so extreme that they are less than the threshold for printing a floating point number are output as 0.

@@

There was no evidence of elevation in risk of  "severe  test positive disease" or "severe test positive disease or death"  in  any non-White ethnic group using  either the ONOMAP  or SMR  approach.

Using the named based classification for which there are much more data  for those of South Asian ethnicity the rate ratio was below 1  and was statistically significant.  A further disaggregation of the name based classification into Muslim South Asian and Non-Muslim South Asian ( data not presented) showed risk ratios of similar size below 1 for both subgroups.  For those of Chinese  ethnicity the name based classification showed lower test positive rates also.

Adjusting for SIMD and residential care home status  did not affect these estimates.

There were increased risk ratios for non-White ethnic groups for test positivity  using SMR  - however note our caveats on interpretation of   this - both because of differential missingness in SMR based ethnicity and because of selection effects on test positive levels.

@@

We do not find evidence at this stage of elevations in risk of severe COVID-19 disease in non-White ethnic groups compared to White.

@@

rownames(table.hospitalized.demog.ethnicsmr) <-
    gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
knitr::kable(table.hospitalized.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized or severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

rownames(table.ethnic5smr) <- gsub("_", " ", rownames(table.ethnic5smr))
knitr::kable(table.ethnic5smr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnic5smr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

@@

@@

rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
knitr::kable(table.hospitalized.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized or severe disease with demographic variables including ethnicity based on name classification") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
knitr::kable(table.severe.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

>>>>

# Background
There have been reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.   For example, a preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95% CI 2.92-5.12); RR 2.11 (95% CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively).

@@

Any additional persons with a registered death with a COVID-19 code (suspected or probable) were also identified.

@@

Data were extracted for this analysis on 8th June 2020. They do not include the Lighthouse data as we do not have access to this through ECOSS as yet.

@@

Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of recording of self-assigned ethnic group in recent years.   Census categories are used. However, there remains substantial missingness.   For all cases and controls the linkage team extracted i) the most recent ethnic group assignation from these two datasets with a 10 year lookback period. Unknown or missing codes were excluded.  Concordance between the most recent SMR ethnic group and the most frequent SMR ethnic group recorded was 98%.

@@

As other sociodemographic factors may vary by ethnicity we further adjusted analyses for residential care home status, Scottish Index of Multiple Deprivation (SIMD) and diabetes.

# Results
SMR based ethnicity was unavailable for approximately 25% of cases and 35% of controls. The ONOMAP approach  assigned ethnicity  in >99% of persons.

@@

Table \ref{tab:crosstab} shows the agreement between the ONOMAP  (name based) assigned ethnicity and the self-assigned SMR ethnicity.  The name classification assigns to the same as  SMR category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[4, 4]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.  The proportion with  non-White ethnicity  is low for all groups. Thus although the sensitivity of ONOMAP for assigning SMR South Asian ethnicity is >99% the positive predictive value of ONOMAP assigned South Asian ethnicity is just 65%.

Clearly analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records. However, the differential missingness between cases and controls means the rate ratios here must be interpreted with caution.  Therefore, in all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.

@@

i)all test positive cases ii) all hospitalized or severe  test positive cases, iii) severe  test positive cases iv) severe test positive cases OR a COVID-19 death registration.  Severe test positive cases are those who  entered a critical care unit  or died within 28 days of a positive test.  Note that full follow up to 28 days has not yet completed for  the recent cases. The first column of rate ratios are univariate. Those in the second column are further adjusted for SIMD and care home status. Tables \ref{tab:ethnicsmrDB} and \ref{tab:ethnicsmrNRSDB} include an additional adjustment for Diabetes.

Subsequent tables show the same comparisons with ethnicity based on ONOMAP name classification. Tables \ref{tab:severity.smr} and \ref{tab:severity} show the distribution over severity categories by ethnicity.  Tables \ref{tab:onomapDB} and \ref{tab:onomapNRSDB} include an additional adjustment for Diabetes. Tables \ref{tab:carehome} and \ref{tab:board} show the distribution of care home status and health board.  Note that  p -values  that are so extreme that they are less than the threshold for printing a floating point number are output as 0.

@@

There is evidence of an elevated risk of  "severe test positive disease" for those of South Asian ethnicity using both the SMR and ONOMAP approach (univariate rate ratios of 1.58 and 1.45 respectively) with borderline significance.  The rate ratios increase and are slightly more significant when SIMD and residential care home status are adjusted for (adjusted rate ratios of 1.68 and 1.59 using SMR and ONOMAP respectively).They are attenuated and lose significance when further adjustment is made for diabetes status (rate ratios of 1.44 and 1.38 using SMR and ONOMAP respectively).

There is also some evidence of an elevation in risk of "severe test positive or death" among those of Black and South Asian ethnicity.  The univariate rate ratio among those of Black ethnicity is 1.12 and increases to 1.38 after adjustment for SIMD and residential care home status.  This effect is observed using the SMR approach since ONOMAP cannot be used to assign Black ethnicity as we describe above. Further adjustment for diabetes status reduces the rate ratio to 1.25.

For those of South Asian ethnicity, the rate ratios associated with "severe test positive or death" using the SMR approach are lower than for "severe test positive disease" and are not significant (univariate rate ratio is 1.23 and rate ratio after adjustment for SIMD and residential care home status is 1.42).  Using the named based classification, for those of South Asian ethnicity the univariate rate ratio is 1.16 and is not significant.  Adjustment for SIMD and residential care home status increases the rate ratio to 1.4.  Further adjustment for diabetes status reduces the rate ratios, using SMR and ONOMAP, to 1.24.

Using the SMR approach, there are significant increased rate ratios for those of Black and South Asian ethnicity for test positivity. However note our caveats on interpretation of this - both because of differential missingness in SMR based ethnicity and because of selection effects on test positive levels.

Significant increased rate ratios for "hospital or severe disease" were observed among those of South Asian ethnicity using both SMR and ONOMAP approaches.

For those of Chinese ethnicity, lower test positivity was observed using the SMR and ONOMAP approaches.  Further to this, using SMR data, there was also a reduced risk of "severe test positive disease" and "severe test positive or death" among this ethnic group.

@@

There are borderline significant elevations in rates of having a positive test, being test-positive and hospitalised or being test-positive and severe (CCU or death) rates in those of South Asian origin.  When deaths with a mention of COVID in those never tested positive are included as fatal cases there is no significant elevation in risk ratio in those of South Asian origin compared to Whites.  Where elevated risk ratios are seen the are increased by adjusting for SIMD and residential care home status so lower socioeconomic status and care home status does not explain these elevations.  Further adjustment for diabetes attenuated the risk ratio consistent with higher prevalence of diabetes contributing to the elevations in risk.

For those of Black ethnic origin, there was evidence of elevation in risk of having a positive test and being test-positive and being hospitalised but not of severe disease or severe disease with additional death registrations in those never tested positive included.

For those of Chinese origin here was evidence of a reduction in risk of having a positive test compared to whites.  Point estimates for hospitalised and severe disease risk ratios were well below 1 but not significantly.  Absolute numbers were very low for this ethnic group.

@@

rownames(table.hospitalized.demog.ethnicsmr) <-
    gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
knitr::kable(table.hospitalized.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized or severe disease among test positives with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

rownames(table.ethnic5smr) <- gsub("_", " ", rownames(table.ethnic5smr))
knitr::kable(table.ethnic5smr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnic5smr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease among test positives with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

```{r ethnic5smrDB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic5smrDB) <- gsub("_", " ", rownames(table.ethnic5smrDB))
knitr::kable(table.ethnic5smrDB,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmrDB",
             col.names=c(clean.header(colnames(table.ethnic5smrDB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease among test positives with ethnicity (additional adjustment for Diabetes) in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

@@

```{r ethnic5smrNRSDB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic5smrNRSDB) <- gsub("_", " ", rownames(table.ethnic5smrNRSDB))
knitr::kable(table.ethnic5smrNRSDB,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmrNRSDB",
             col.names=c(clean.header(colnames(table.ethnic5smrNRSDB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease OR a COVID-19 death with ethnicity (additional adjustment for Diabetes) in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

@@

rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
knitr::kable(table.hospitalized.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized or severe disease among test positives with demographic variables including ethnicity based on name classification") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
knitr::kable(table.severe.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease among test positives with demographic variables including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```{r onomapDB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.severe.demogDB) <- gsub("_", " ", rownames(table.severe.demogDB))
knitr::kable(table.severe.demogDB,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomapDB",
             col.names=c(clean.header(colnames(table.severe.demogDB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease among test positives with demographic variables (additional adjustment for Diabetes) including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

@@

```{r onomapNRSDB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.severe.demogNRSDB) <- gsub("_", " ", rownames(table.severe.demogNRSDB))
knitr::kable(table.severe.demogNRSDB,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomapNRSDB",
             col.names=c(clean.header(colnames(table.severe.demogNRSDB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease OR any COVID-19 death with demographic variables (additional adjustment for Diabetes) including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")
